Astrophysical calibration of gravitational-wave detectors

### 2015-07-16 docs/cosmic_cal.tex
Written some more results

% acronym definitions
\acrodef{GW}[GW]{gravitational wave}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{MWEG}[MWEG]{Milky Way equivalent galaxies}
\acrodef{BBH}[BBH]{binary black hole}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{NSBH}[NSBH]{neutron star-black hole}
\acrodef{SNR}[SNR]{signal-to-noise ratio}
\acrodef{GRB}[GRB]{gamma-ray burst}
\acrodef{sGRB}[sGRB]{short gamma-ray burst}
\acrodef{CBC}[CBC]{compact binary coalescence}
\acrodef{EM}[EM]{electro-magnetic}

@@

The technique used throughout GW research in calibrating the detectors is
carried out through a complicated system of physical manipulations of the
Fabret-Perot Michelson interferometer; where an elaborate feedback system is
used to sustain a defined measurement in arm length difference between the
moving mirrors~\cite{LIGOCal}. In general, the interferometers are set to move
at a frequency corresponding to a GW signal. An error signal is then found
through the Pound-Driver-Hall technique~\cite{Black} for mirror stabilisation
which is proportional to the difference in mirror separation of the
Fabret-Perot Michelson interferometer. The Pound-Drever-Hall technique is
essentially a control loop feedback system in which the difference output (in
GW detectors, this is the distance the mirrors move due to the GW) and the the
output set is reduced to the highest degree of certainty. This is just a brief
overview of the calibration method currently used, and is an extensive study in
itself, see \cite{Vitale:2012} and references therein. For advanced LIGO and Virgo,
at the frequency range used in this project, the error in the calibration using
the `hardware injection method' is roughly $10\%$ \cite{Vitale:2012}. This is a
benchmark set for the error estimation using the proposed new method in this
project.~\cm{Clean this up and expand.}

@@

For a non-spinning system the general form of the frequency-domain polarisation amplitudes are
%
\begin{eqnarray}\label{eq:signal}
  \gws_{+}(f) &\propto \frac{1+\cos^{2}\iota}{D_{L}}
\mathcal{M}^{5/6}f^{-7/6}e^{-i\Psi(f, \nu, t_\mathrm{c}, \phi_{\mathrm{c}})} \nonumber \\
  \gws_{\times}(f) &\propto
 \frac{\cos^{2}\iota}{D_{L}}\mathcal{M}^{5/6} f^{-7/6}e^{-i\Psi(f, \nu, t_\mathrm{c}, 
\phi_{\mathrm{c}})}
\end{eqnarray}
%
where the chirp-mass $\mathcal{M}$ is defined as $\mathcal{M}=M\eta^{3/5}$, with 
$\eta=m_{1}m_{2}/M^{2}$ and $M=m_{1}+m_{2}$, $\iota$ is the inclination angle, and $\nu=(\pi 
Mf)^{1/3}$.  The \ac{GW} strain measured at the $k^{\textrm{th}}$ detector is then given by
%
\begin{equation}
  \label{eq:gravsig}
   \gws^{k}(f) = \big[ F_{+}^{k}(\alpha, \delta, \psi)\gws_{+}(f) +
F_{\times}^D(\alpha, \delta, \psi)\gws_{\times}(f)\big]
\end{equation}
%
where $F_{+},F_{\times}$ are the antenna response functions which are dependent
upon the polarisation angle $\psi$ and the sky position of the source
$\alpha,\delta$.

@@

To estimate how well we can constrain the calibration scaling each detector in an advanced 
detector network we have performed simulations with injected \ac{BNS} and \ac{NSBH} signals at a 
range of distances (50\,Mpc to 500\,Mpc with 50\,Mpc increments for the \ac{BNS} signals and 
100\,Mpc to 1000\,Mpc with 100\,Mpc increments for the \ac{NSBH} signals). The simulations use the 
two Advanced LIGO detectors (one at Hanford -- H1 -- and one at Livingston -- L1) and the Advanced 
Virgo detector (V1). The simulations are all performed in the freqeuncy domain and span a frequency 
range from 20\,Hz to 400\,Hz. We assume all detectors are expected to be operating at their design
sensitivity.

% Describe the ensemble of simulations used
As described in Section~\ref{sec:priors} we had two different scenarios for the prior distribution 
of source system inclination angles, so we have produced separate simulations for both cases. In 
each case 1000 sources were created with distributions on the inclination, component masses,
reference phase and polarisation angle drawn from their respective prior distributions. For the 
time of coalescence the times were held fixed within the centre of the prior window. When drawing 
the sky position points were drawn uniformly over the sky. However, with all these an additional 
constraint was imposed based on the fact that we require a distributions sources that could 
actually have been detected. We therefore include the requirement from \cite{2012PhRvD..85h2002A} 
that their be at least two detectors in the network for which the SNR is $\gtrsim 5.5$ (as these 
would be signals coincident with a GRB we do not include the often used further constraint that 
the total coherent network SNR is greater than 12). This generally means that as we go to larger
distances our population of sources that fulfil this criterion will not be uniformly distributed
over the sky, and will generally be closer to being circularly polarised.

For each simulation the calibration scale factor for each detector has been drawn from a Gaussian 
distribution with a mean of unity and a standard deviation of 0.125 (equivalent to the detectors 
having on average a $\pm 10\%$ amplitude calibration error).

@@

For each of the simulated sources we have calculated the 68\% credible region on the scale factors 
for each detector (for a Gaussian distribution this is equivalent to the region bounded by 
$1\sigma$ intervals). For all the signals in each distance increment we have produced the 
distribution of the half widths (i.e.\ $1\sigma$) of these scale factors. These are shown in 
figures~\ref{fig:narrowdist} and \ref{fig:widedist} for the case with the inclination distribution 
with a width of $5^{\circ}$ and $20^{\circ}$ respectively.

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factors_5degs_boxplot.png}
 \end{center}
 \caption{\label{fig:narrowdist} Add caption!}
\end{figure}

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factors_20degs_boxplot.png}
 \end{center}
 \caption{\label{fig:widedist} Add caption!}
\end{figure}

@@

The main results obtained in this project were from analysing randomly chosen
sky positions from the entire field of view for the detector. Through these
simulations, it was shown that this new method of calibration can reach and
surpass the current method in terms of accuracy. The results in
figure~\ref{fig:std-rndsp} show that for half of the randomly chosen sky
positions relate to an estimation of $s$ within a range comparable with the
current methods.

>>>>

% acronym definitions
\acrodef{GW}[GW]{gravitational wave}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{MWEG}[MWEG]{Milky Way equivalent galaxies}
\acrodef{BBH}[BBH]{binary black hole}
\acrodef{BNS}[BNS]{binary neutron star}
\acrodef{NSBH}[NSBH]{neutron star-black hole}
\acrodef{SNR}[SNR]{signal-to-noise ratio}
\acrodef{GRB}[GRB]{gamma-ray burst}
\acrodef{sGRB}[sGRB]{short gamma-ray burst}
\acrodef{CBC}[CBC]{compact binary coalescence}
\acrodef{EM}[EM]{electro-magnetic}
\acrodef{aLIGO}[aLIGO]{Advanced LIGO}
\acrodef{AdV}[AdV]{Advanced Virgo}

@@

The technique used throughout GW research in calibrating the detectors is
carried out through a complicated system of physical manipulations of the
Fabret-Perot Michelson interferometer; where an elaborate feedback system is
used to sustain a defined measurement in arm length difference between the
moving mirrors~\cite{LIGOCal}. In general, the interferometers are set to move
at a frequency corresponding to a GW signal. An error signal is then found
through the Pound-Driver-Hall technique~\cite{Black} for mirror stabilisation
which is proportional to the difference in mirror separation of the
Fabret-Perot Michelson interferometer. The Pound-Drever-Hall technique is
essentially a control loop feedback system in which the difference output (in
GW detectors, this is the distance the mirrors move due to the GW) and the the
output set is reduced to the highest degree of certainty. This is just a brief
overview of the calibration method currently used, and is an extensive study in
itself, see \cite{Vitale:2012} and references therein. For \ac{aLIGO} and \ac{AdV},
at the frequency range used in this project, the error in the calibration using
the `hardware injection method' is roughly $10\%$ \cite{Vitale:2012}. This is a
benchmark set for the error estimation using the proposed new method in this
project.~\cm{Clean this up and expand.}

@@

For a non-spinning system the general form of the frequency-domain polarisation amplitudes are
%
\begin{eqnarray}\label{eq:signal}
  \gws_{+}(f) &\propto \frac{1+\cos^{2}\iota}{D_{L}}
\mathcal{M}^{5/6}f^{-7/6}e^{-i\Psi(f, \nu, t_\mathrm{c}, \phi_{\mathrm{c}})} \nonumber \\
  \gws_{\times}(f) &\propto
 \frac{\cos^{2}\iota}{D_{L}}\mathcal{M}^{5/6} f^{-7/6}e^{-i\Psi(f, \nu, t_\mathrm{c}, 
\phi_{\mathrm{c}})}
\end{eqnarray}
%
where the chirp-mass $\mathcal{M}$ is defined as $\mathcal{M}=M\eta^{3/5}$, with 
$\eta=m_{1}m_{2}/M^{2}$ and $M=m_{1}+m_{2}$, $\iota$ is the inclination angle, and $\nu=(\pi 
Mf)^{1/3}$.  The \ac{GW} strain measured at the $k^{\rm{th}}$ detector is then given by
%
\begin{equation}
  \label{eq:gravsig}
   \gws^{k}(f) = \big[ F_{+}^{k}(\alpha, \delta, \psi)\gws_{+}(f) +
F_{\times}^D(\alpha, \delta, \psi)\gws_{\times}(f)\big]
\end{equation}
%
where $F_{+},F_{\times}$ are the antenna response functions which are dependent
upon the polarisation angle $\psi$ and the sky position of the source
$\alpha,\delta$.

@@

To estimate how well we can constrain the calibration scaling for each detector in an advanced 
detector network we have performed simulations with injected \ac{BNS} and \ac{NSBH} signals at a 
range of distances (50\,Mpc to 500\,Mpc with 50\,Mpc increments for the \ac{BNS} signals and 
100\,Mpc to 1000\,Mpc with 100\,Mpc increments for the \ac{NSBH} signals). The simulations use the 
two Advanced LIGO detectors (one at Hanford -- H1 -- and one at Livingston -- L1) and the Advanced 
Virgo detector (V1). The simulations are all performed in the freqeuncy domain and span a frequency 
range from 20\,Hz to 400\,Hz. We assume all detectors are expected to be operating at their design
sensitivity (as given by Figure 1 of \cite{2013arXiv1304.0670L}). As such we add noise to each 
simulation based on the sensitivity curve, but scaled with the associated calibration scale factor.

% Describe the ensemble of simulations used
Separately for the \ac{BNS} and \ac{NSBH} systems we have performed 2000 simulations at each 
distance value. For proposing injections for each simulation the source parameters were randomly 
drawn from the prior distributions given in Section~\ref{sec:priors}, with the sky position being 
drawn randomly from a uniform distribution, and with the scaling factors for each detector drawn 
from a Gaussian distribution with a mean of one and standard deviation of $0.125$ (equivalent to a 
mean fractional offset of 10\%). However, in accepting a proposed injection as one to be analysed 
we introduced a criteria that the signal be ``detectable'', based on it having an SNR of $\geq 5.5$ 
in at least two detectors as used in \cite{2012PhRvD..85h2002A} (as these would be signals 
coincident with a GRB we do not include the often used further constraint that the total coherent 
network SNR is greater than 12). The coalescence times for all injections were held fixed 
within the centre of the prior window. This criteria means that as we go to larger distances our 
population of injected sources will not be uniformly distributed over the sky, and will generally 
be closer to being circularly polarised. For the \ac{NSBH} systems it also means that at larger 
distances we will have more sources with larger mass ratios \MP{Check this}.

@@

For each of the simulated sources we have used the {\tt emcee} python MCMC package to perform 
parameter estimation over the unknown source parameters using the prior as discussed in 
Section~\ref{sec:priors}. In each case we use an estimate of the noise PSD based on the design 
sensitivity, but simulated as if estimated by averaging 64 separate PSD estimates and scaled with 
the same factor as applied to the injection and noise\footnote{We do not account for there being 
a difference between the estimated PSD and the actual PSD of the analysed section of data. This 
difference would be very much correlated with the scale factor, so in reality our estimate of the 
scale factor would be a combination of the calibration offset and the difference in the PSD. As 
such our results on real data would be an upper limit on the calibration scale factor.}. This has 
provided posterior probability distributions on the calibration scale factors for each detector 
\MP{Show an example for both source types}. From these we have calculated the 68\% credible region 
on the scale factors for each detector (if these were Gaussian distributions this is equivalent to 
the region bounded by $1\sigma$ intervals). For all the signals at each distance increment we have 
produced the distribution of the fractional half widths (i.e.\ $1\sigma$) of these scale factors 
confidence intervals compared to the actual value. These are shown as boxplots in 
figures~\ref{fig:bnsresults} and \ref{fig:nsbhresults} for the \ac{BNS} and \ac{NSBH} systems 
respectively. The boxes show the extent from the lower to upper quartile of the values, whilst the 
whiskers represent range of values. The black line within each box gives the median value and the 
star gives the mean value. Also shown on each plot as the dashed magenta line is the percentage of 
proposed injections that actually fulfilled the detection criterion.

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factor.pdf}
 \end{center}
 \caption{\label{fig:bnsresults} Distribution of the percentage accuracy at which the calibration 
scale factor can be determined for a three detector network if assuming \ac{BNS} systems (provided 
a coincident GRB is observed and can yield a distance estimate). The boxplots span the lower to 
upper quartile range of the distributions, with the median value shown as a horizontal line within 
the box and the mean shown as a star. The dashed magenta line shows the percentage of source drawn 
from the prior distribution that would be detectable at each distance value.}
\end{figure}

\begin{figure}
 \begin{center}
  \includegraphics[width=1.0\textwidth]{scale_factor_nsbh.pdf}
 \end{center}
 \caption{\label{fig:nsbhresults} Distribution of the percentage accuracy at which the calibration 
scale factor can be determined for a three detector network if assuming \ac{NSBH} systems (provided 
a coincident GRB is observed and can yield a distance estimate). The plot contents are the same as 
in figure~\ref{fig:bnsresults}.}
\end{figure}

In figure~\ref{fig:bnsresults} we see that unsurprisingly on average the scale factor can be 
recovered to equivalent precision in both \ac{aLIGO} detectors, with uncertainties generally within 
the 10\% range for sources at 100\,Mpc. An interesting feature that we see is that for distances 
$\gtrsim 250$\,Mpc the upper extent of the uncertainty hits a maximum at $\sim 25\mbox{--}30\%$, 
whilst the width of boxes narrows. This is due to our ``detectability'' criteria, in that for all 
distances we will only be seeing those sources with high enough SNR that we would consider them 
detectable. There will therefore be fewer sources with SNR higher than this criteria at large 
distances, giving us a narrower range, and we automatically exclude those with SNR that are too 
small thus truncating our uncertainty distribution at the upper range. However, this does show that 
on average for sources that are detectable out to 450\,Mpc we would be able to constrain the 
calibration scale factor uncertainty to $\sim 20\%$. As the detection criteria will generally come 
from the SNR being high enough within the two \ac{aLIGO} instruments this means that the SNR in 
\ac{AdV} can be small and thus the ability to constrain the scale factor for it becomes poor 
(although it still provides information that the calibration is not grossly inaccurate).

@@

The main results obtained in this project were from analysing randomly chosen
sky positions from the entire field of view for the detector. Through these
simulations, it was shown that this new method of calibration can reach and
surpass the current method in terms of accuracy.
